---
title: "Phase3"
author: "Karla Otoude and Chloé Marquis"
format: pdf
execute:
  echo: false
---

# Link to our github project 
[Our Github Project](https://github.com/chloemrq/Project-Data-Management)

# Research question 
**How does the level of social protection impact the general population health ?** 

We want to identify if a stronger social and health protection (quantified by different variables such as health expenditure, tax revenues, etc) has an impact on the avoidable mortality rate. In order to do so we will analyse data collected in OECD countries from 2000 to 2023. 

There are two goals in our analysis :
The first is to understand the global evolution of the data and the reasons behind its trends.
The second is to study the Covid-19 crisis. We believe it revealed how necessary having a proper social protection is to limit the number of deaths during a pandemic. 


# Method 
To conduct our study we will be using a **linear mixed model (LMM)** with lags. Such a model will allow us to properly study the OECD countries taking into account their differences. It also a fitting model since we are working with time-based data. 

Considering our research question, we will use multiple datasets and variables.

For explanatory variables, in order to estimate the strength and quality of a country’s health protection scheme, we will have : 
- Health expenditure 
- Social expenditure aggregates
- Number of hospitals beds 

We also have more general variables such as :
- GDP per capita 
- Unemployment rate 
- Human development index 
- Life expectancy
They will be used as control variables, since richer countries tend to have better living conditions which may lead to a generally healthier population, without the need for a strong health protection. 

Finally, the explained variable will be : 
- Avoidable mortality 
It will be used to estimate the population’s general health. 

# Loading the packages 
```{r}
#| message: false
library(here)
here::i_am("Project-Data-Management.Rproj")
```


```{r}
#| message: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(vroom)
library(readxl)
```


# Loading the cleaned dataset from phase 2
```{r}
database <- vroom(here("data", "cleaned_database.csv"))
```








